---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


# Introduction

With the advancement of technology and the growth of the **big data** movement, it is now possible to collect a large amount of data about personal movement using activity monitoring devices.  Such examples are: [Fitbit](http://www.fitbit.com/), [Nike Fuelband](http://www.nike.com/us/en_us/c/nikeplus-fuelband), or [Jawbone Up](https://jawbone.com/up). These kinds of devices are part of the "quantified self" movement: those who  measurements about themselves on a regular basis in order to improve their health,  find patterns in their behaviour, or because they are simply technology geeks. However, these data remain severely underused due to the fact that the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals throughout the day. The data consists of two months of data from an anonymous individual collected during the months of October and November 2012 and include the number of steps taken in 5 minute intervals each day.

The overall goal of this assignment is to make some basic exploratory data analysis to assess some activity patterns with regards to the anonymous individual's walking patterns.  For each day, there are readings taken at particular 5-minute intervals.  These readings correspond to the number of **steps** taken by the anonymous individual between the previous 5-minute interval to the current 5-minute interval.

# Data Layout

These intervals are assigned a unique ID which corresponds to the interval taken at that day.  As an example, the ID of `15` means that at the third observation (i.e. at the 15 minute mark), that is when the reading was taken for this day.  There are 61 days between October and November, and as such there are `61` observations recorded at the 15 minute mark.  Furthermore, there are 61 observations taken at the same time interval as they were taken each day within these two months.  As such, there are 61 observations for the 20 minute mark, the 30 minute mark, the 45 minute mark and so on.  In addition, for each day, there are `288` intervals / observations were recorded at per day.

As such, the data provided to us is a 3-column data frame:

1.  Column 1 - *steps*: Indicates the number of steps taken during a neighbouring 5-minute interval (between 5 minutes and 10 minutes, 10 minutes and 15 minutes, etc.).  There were some instances where there were no readings taken, most likely due to the fact that the subject was sleeping or the device was shut off.  These are coded in as `NA` values.
2.  Column 2 - *date*:  Indicates the date at which the measurement was taken.  This is in the `YYYY-MM-DD` format.
3.  Column 3 - *interval*: The aforementioned ID that determines at which 5-minute interval the reading was taken at (5, 10, 15, 20, etc.)

# Procedure

## Preamble

There are five steps overall in our analysis:

1.  Loading in and preprocessing the data
2.  Plotting a histogram of the total number of steps taken each day and calculating the mean and median of each day.
3.  Determining the average daily activity pattern: Plotting a time-series plot for each 5-minute interval (the x-axis) with the average number of steps averaged across all days (the y-axis).  This means that for each 5-minute interval, calculate the average of the 61 observations taken at each interval.  We also calculate the median of the 61 observations taken at each interval as well.
4.  Imputing missing values: There were some instances where the amount of steps read for an interval were missing.  These were coded in as `NA` values.  The presence of missing days may introduce bias into some calculations or summaries of the data.  In this step, a simple strategy was performed to replace the missing values in the dataset with a filler value.  This value was chosen to be the **mean within the 5-minute interval the value was missing for**.  For example, suppose on October 5, 2012, at the 10-minute interval there is missing data.  This data is filled in by the mean of whatever data was available overall (the 61 observations excluding the `NA` values) for this particular interval.  The analysis of Step #2 is repeated with similarities and differences being reported.
5.  The last part of the analysis is to split up the data into weekdays and weekends and observe if there is any difference in activity patterns between these two classes.  The same kind of plot is repeated like in Step #3, but now there are two separate plots to reflect the activity on the weekdays and weekends.

## Loading in and preprocessing data

We assume that the reader set the correct R working directory with the setwd() function.

Load the data (i.e. read.csv())

```{r}
# Clear the workspace
rm(list=ls())

# Load the raw activity data
activity_raw <- read.csv("activity.csv", stringsAsFactors=FALSE)
```


Process/transform the data (if necessary) into a format suitable for analysis

```{r}
# Load the raw activity data
activity_raw <- read.csv("activity.csv", stringsAsFactors=FALSE)
```

Process/transform the data (if necessary) into a format suitable for analysis

```{r}

# Transform the date attribute to an actual date format
activity_raw$date <- as.POSIXct(activity_raw$date, format="%Y-%m-%d")

# Compute the weekdays from the date attribute
activity_raw <- data.frame(date=activity_raw$date, 
                           weekday=tolower(weekdays(activity_raw$date)), 
                           steps=activity_raw$steps, 
                           interval=activity_raw$interval)

# Compute the day type (weekend or weekday)
activity_raw <- cbind(activity_raw, 
                      daytype=ifelse(activity_raw$weekday == "sabato" | 
                                     activity_raw$weekday == "domenica", "weekend", 
                                     "weekday"))

# Create the final data.frame
activity <- data.frame(date=activity_raw$date, 
                       weekday=activity_raw$weekday, 
                       daytype=activity_raw$daytype, 
                       interval=activity_raw$interval,
                       steps=activity_raw$steps)

# Clear the workspace
rm(activity_raw)


```

What is the mean total number of steps taken per day?
For this part of the assignment, you can ignore the missing values in the dataset.

1 - Make a histogram of the total number of steps taken each day

```{r}
# Compute the total number of steps each day (NA values removed)
sum_data <- aggregate(activity$steps, by=list(activity$date), FUN=sum, na.rm=TRUE)

# Rename the attributes
names(sum_data) <- c("date", "total")
```

The histogram is given by the following lines of code:

```{r}
# Compute the histogram of the total number of steps each day
hist(sum_data$total, 
     breaks=seq(from=0, to=25000, by=2500),
     col="blue", 
     xlab="Total number of steps", 
     ylim=c(0, 20), 
     main="Histogram of the total number of steps taken each day\n(NA removed)")
```

2 - Calculate and report the mean and median total number of steps taken per day

```{r}
mean(sum_data$total)
median(sum_data$total)
```


## What is the average daily activity pattern?

1 - Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
# Clear the workspace
rm(sum_data)

# Compute the means of steps accross all days for each interval
mean_data <- aggregate(activity$steps, 
                       by=list(activity$interval), 
                       FUN=mean, 
                       na.rm=TRUE)

# Rename the attributes
names(mean_data) <- c("interval", "mean")
```

```{r}
head(mean_data)
```

The time serie plot is created by the following lines of code

```{r}

# Compute the time series plot
plot(mean_data$interval, 
     mean_data$mean, 
     type="l", 
     col="blue", 
     lwd=2, 
     xlab="Interval [minutes]", 
     ylab="Average number of steps", 
     main="Time-series of the average number of steps per intervals\n(NA removed)")

```

2 - Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
# We find the position of the maximum mean
max_pos <- which(mean_data$mean == max(mean_data$mean))

# We lookup the value of interval at this position
max_interval <- mean_data[max_pos, 1]

# Clear the workspace
rm(max_pos, mean_data)
```

```{r}
max_interval
```

## Inputing the missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

1 - Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NA's)


```{r}
# Clear the workspace
rm(max_interval)

# We use the trick that a TRUE boolean value is equivalent to 1 and a FALSE to 0.
NA_count <- sum(is.na(activity$steps))
```

```{r}
NA_count
```

2 - Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

```{r}
# Clear the workspace
rm(NA_count)

# Find the NA positions
na_pos <- which(is.na(activity$steps))

# Create a vector of means
mean_vec <- rep(mean(activity$steps, na.rm=TRUE), times=length(na_pos))
```

We replace each NA value with the mean of the steps attribute.

3 - Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
# Replace the NAs by the means
activity[na_pos, "steps"] <- mean_vec

# Clear the workspace
rm(mean_vec, na_pos)
```

```{r}
head(activity)
```

4 - Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
# Compute the total number of steps each day (NA values removed)
sum_data <- aggregate(activity$steps, by=list(activity$date), FUN=sum)

# Rename the attributes
names(sum_data) <- c("date", "total")

# Compute the histogram of the total number of steps each day
hist(sum_data$total, 
     breaks=seq(from=0, to=25000, by=2500),
     col="blue", 
     xlab="Total number of steps", 
     ylim=c(0, 30), 
     main="Histogram of the total number of steps taken each day\n(NA replaced by mean value)")
```

```{r}
mean(sum_data$total)
```

```{r}
median(sum_data$total)
```

These values differ greatly from the estimates from the first part of the assignment. The impact of imputing the missing values is to have more data, hence to obtain a bigger mean and median value.

##Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

1 - Create a new factor variable in the dataset with two levels - "weekdays" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
# The new factor variable "daytype" was already in the activity data frame
head(activity)
```

2 - Make a panel plot containing a time series plot (i.e. type = "l") of the 5- minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
# Clear the workspace
rm(sum_data)

# Load the lattice graphical library
library(lattice)

# Compute the average number of steps taken, averaged across all daytype variable
mean_data <- aggregate(activity$steps, 
                       by=list(activity$daytype, 
                               activity$weekday, activity$interval), mean)

# Rename the attributes
names(mean_data) <- c("daytype", "weekday", "interval", "mean")
```

```{r}
head(mean_data)
```

The time series plot take the following form:

```{r}
# Compute the time serie plot
xyplot(mean ~ interval | daytype, mean_data, 
       type="l", 
       lwd=1, 
       xlab="Interval", 
       ylab="Number of steps", 
       layout=c(1,2))
```


